---
title: "Simple Example (the hard way)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simple Example (the hard way)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}
library(tibble)
library(dplyr)
library(wext)
```

The data used for this example is the same data set in the "simple" example from the Raphael group's [GitHub repository](https://github.com/raphael-group/wext/tree/master/examples/simple). The original file format was one line per sample with a list of the genes that were mutated. I have already prepared this into a tidy format in "data-raw/simple_dataset.R", as it is likely what the user will start with.

```{r}
simple_dataset
```

In this vignette, I show how WExT works by running each step separately. The ["Simple Example (the easy way)"]() explains how a normal user would run the analysis.

---

## Building the weighted matrix for the row-exclusivity test $W_R$.

This is simply defined as

$$
W_R = \frac{1}{\Omega_R} \sum_{B \in \Omega_R}{B}
$$

but since every row of $B \in \Omega_R$ can be considered independently, this simplifies to

$$
W_R = \left[ w_{ij} = \frac{r_i}{n} \right]
$$

Thus, we compute the weights for the data

```{r calc_WR}
W_R <- calculate_row_exclusivity_weights(dat = simple_dataset,
                                         sample_col = sample_name,
                                         mutgene_col = mutated_gene)
```

```{r show_WR, echo = FALSE}
W_R
```


